---
title: "MA Head Start Disability Data Brief"
subtitle: "Data Ethnography and Advocacy Lab, 2026"
author: "FILL"
format: 
  revealjs:
    scrollable: true
    df-print: paged
    theme: [default]
    incremental: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 5
editor: source
execute:
  echo: false
  messages: false
  warnings: false
---

```{r}
library(tidyverse)

source("pir_source.R")
source("disability_source.R")
```

# Headline Statistics

```{r}
hs_2024 <- disability_under_5 |>
  filter(ST == "MA", YEAR_END == 2024)

hs_2024 <- disability_under_5 |>
  filter(ST == "MA", YEAR_END == 2024)

# 1. Statewide total disabled 0-5
hs_2024 |> pull(IDEA_DISABLED_0_5)

# 2. Statewide proportion of 0-5 disabled
hs_2024 |> pull(STATEWIDE_PROP_DISABLED)

# 3. Head Start served, disabled (IEP + IFSP)
hs_2024 |> pull(HS_DISABLED)

# 4. Share of statewide disabled covered by Head Start
hs_2024 |> pull(HS_SHARE_OF_STATE)

# 5. % of Head Start children who are disabled
hs_2024 |> pull(PERCENT_HS_DISABLED)
```


## Disability Enrollment in MA vs Other States

```{r}
combined_pir |>
  filter(pir_year == 2025) |>
  group_by(State) |>
  summarize(total_child_enroll = sum(A.10.g), 
      total_disability_enroll = sum(C.22.a.1),
      percent_disability = total_disability_enroll / total_child_enroll * 100) |>
  arrange(desc(percent_disability))
```

## Tracking Indicator 11

```{r}
# The number who received an evaluation to determine IDEA eligibility / The total number of children referred for an evaluation to determine eligibility under the Individuals with Disabilities Education Act (IDEA) during the program year

combined_pir |> 
  filter(pir_year == 2025, State == "MA") |> 
  summarize(ratio = sum(C.22.a) / sum(C.22)) |> 
  pull(ratio)
```

Indicator 11 from the Massachusetts Department of Elementary and Secondary Education is defined as the number of students with initial evaluation assessments completed within 30 school days divided by the number of students for whom parental consent for initial evaluations was received. From the years 2019-2023, Indicator 11 across the State were as follows:
- FFY2019 — School Year 2019-2020: 97.28%
- FFY2020 — School Year 2020-2021: 93.49%
- FFY2021 — School Year 2021-2022: 88.38%
- FFY2022 — School Year 2022-2023: 95.32%
- FFY2023 — School Year 2023-2024: 85.75%

Across Head Start programs, a somewhat analogous statistic on the number who received an evaluation to determine IDEA eligibility divided by the total number of children referred for an evaluation to determine eligibility under the Individuals with Disabilities Education Act (IDEA) during the program year was only 68.39%.

# Evaluation "Waitlist" (Pending Evaluations)

C.23.c counts children referred for an IDEA evaluation whose evaluation was still pending and not yet completed by the responsible agency at year end.

## Pending Share of Referrals Over Time

### All Head Start Programs

```{r}
options(scipen = 999)
pir_c |>
  filter(State == "MA") |>
  group_by(pir_year) |>
  summarize(referred = sum(C.22),
            pending = sum(C.23.c),
            percent_pending = pending / referred * 100) |>
  ggplot(aes(x = pir_year, y = percent_pending)) +
  geom_line() +
  scale_x_continuous(breaks = min(pir_c$pir_year):max(pir_c$pir_year)) +
  scale_y_continuous(limits = c(0, 30)) +
  labs(x = "Year",
       y = "% of Referred Children With Evaluation Pending",
       title = "Head Start PIR, MA, 2021-2025") +
  theme_minimal()
```

### By Program

```{r}
pir_c |>
  filter(State == "MA") |>
  group_by(pir_year, `Type`) |>
  summarize(referred = sum(C.22),
            pending = sum(C.23.c),
            percent_pending = pending / referred * 100) |>
  ggplot(aes(x = pir_year, y = percent_pending)) +
  geom_line() +
  facet_wrap(vars(`Type`)) +
  scale_x_continuous(breaks = min(pir_c$pir_year):max(pir_c$pir_year)) +
  labs(x = "Year",
       y = "% of Referred Children With Evaluation Pending",
       title = "Head Start PIR, MA, 2021-2025") +
  theme_minimal()
```

## Pending as a Share of Non-Evaluations/Reasons for Non-Evaluation

From 2023-2025 pending "Pending" has consistently been the largest single reason a referral goes unevaluated, but its share of non-evaluations fell from  roughly 64% in 2023 to 44% in 2025.

```{r}
pir_c |>
  filter(State == "MA") |>
  group_by(pir_year) |>
  summarize("RTI assignment (C.23.a)" = sum(C.23.a),
            "Parent refused (C.23.b)" = sum(C.23.b),
            "Pending (C.23.c)" = sum(C.23.c),
            "Other (C.23.d)" = sum(C.23.d),
            "Unexplained" = sum(C.22.b) - sum(C.23.a + C.23.b + C.23.c + C.23.d),
            "Total Not Evaluated (C.22.b)" = sum(C.22.b)) |>
  pivot_longer(-pir_year,
               names_to = "Reason for non-evaluation",
               values_to = "n") |>
  pivot_wider(names_from = pir_year, values_from = n)
```

# Children with IEPs/IFSPs Not Receiving Services

### Children with IEPs

```{r}
options(scipen = 999)
pir_c |>
  filter(State == "MA") |>
  group_by(pir_year) |>
  summarize(not_receiving_IEP = sum(C.24.b, na.rm = TRUE)) |>
  ggplot(aes(x = pir_year, y = not_receiving_IEP)) +
  geom_line() +
  scale_x_continuous(breaks = min(pir_c$pir_year):max(pir_c$pir_year)) +
  scale_y_continuous(limits = c(0, 45)) +
  labs(x = "Year",
       y = "# Children with IEP Not Receiving Services",
       title = "Head Start PIR, MA, 2023-2025") +
  theme_minimal()
```

### Children with IFSPs

```{r}
options(scipen = 999)
pir_c |>
  filter(State == "MA") |>
  group_by(pir_year) |>
  summarize(not_receiving_IEP = sum(C.25.b, na.rm = TRUE)) |>
  ggplot(aes(x = pir_year, y = not_receiving_IEP)) +
  geom_line() +
  scale_x_continuous(breaks = min(pir_c$pir_year):max(pir_c$pir_year)) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(x = "Year",
       y = "# Children with IEP Not Receiving Services",
       title = "Head Start PIR, MA, 2023-2025") +
  theme_minimal()
```



