---
title: "MA Head Start Homelessness Data Brief"
subtitle: "Data Ethnography and Advocacy Lab, 2026"
author: "FILL"
format: 
  revealjs:
    scrollable: true
    df-print: paged
    theme: [default]
    incremental: true
    chalkboard:
      theme: whiteboard
      boardmarker-width: 5
editor: source
execute:
  echo: false
  messages: false
  warnings: false
---


```{r}
library(tidyverse)
library(ggrepel)

source("pir_source.R")
source("homeless_source.R")
```

```{r}
options(scipen = 999)
combined_pir |>
  mutate(is_ma = case_when(State == "MA" ~ "MA",
                           State != "MA" ~ "Other States")) |>
  group_by(pir_year, is_ma) |>
  filter(!is.na(is_ma)) |>
  summarize(homeless_children = sum(C.48, na.rm = TRUE),
            total_child_enroll = sum(A.10.g, na.rm = TRUE),
      percent_homeless = homeless_children / total_child_enroll * 100) |>
  ggplot(aes(x = pir_year, y = percent_homeless, col = is_ma, group = is_ma)) +
  geom_line() +
  scale_x_continuous(breaks = min(combined_pir$pir_year):max(combined_pir$pir_year)) +
  scale_y_continuous() +
  labs(x = "Year", 
       y = "% Enrolled Homeless Children",
       col = "",
       title = paste("Head Start PIR, MA,",
                     min(combined_pir$pir_year), 
                     "-" , 
                     max(combined_pir$pir_year))) +
  theme_minimal()  +
  geom_text(aes(label = round(percent_homeless, 2)), 
            vjust = -0.25)

```

---

```{r}
combined_pir |>
  filter(State == "MA") |>
  group_by(pir_year, `Type`) |>
  summarize(homeless_children = sum(C.48),
            total_child_enroll = sum(A.10.g),
      percent_homeless = homeless_children / total_child_enroll * 100) |>
  ggplot(aes(x = pir_year, y = percent_homeless)) +
  geom_line() +
  facet_wrap(vars(`Type`)) +
  scale_x_continuous(breaks = min(combined_pir$pir_year):max(combined_pir$pir_year)) +
  labs(x = "Year", 
       y = "% Enrolled Homeless Children", 
       title = paste("Head Start PIR, MA,",
                     min(combined_pir$pir_year), 
                     "-" , 
                     max(combined_pir$pir_year))) +
  theme_minimal()  +
  geom_text(aes(label = round(percent_homeless, 2)), 
            vjust = -0.25,
            size = 3)
```


---

```{r}
percent_homeless_table <-
  combined_pir |>
  filter(pir_year == 2025) |>
  group_by(State) |>
  summarize(total_child_enroll = sum(A.10.g, na.rm = TRUE), 
      total_homeless_enroll = sum(C.48, na.rm = TRUE),
      num_homeless_na = sum(is.na(C.48)),
      percent_homeless = total_homeless_enroll / total_child_enroll * 100) |>
  arrange(desc(percent_homeless)) |>
  head(10)

percent_homeless_table |>
  select(State, percent_homeless)

```

---

```{r}
percent_homeless_table |>
  mutate(total_not_homeless_enroll = total_child_enroll - total_homeless_enroll) |>
  pivot_longer(cols = c(total_homeless_enroll, total_not_homeless_enroll),
               names_to = "homeless_flag",
               values_to = "value") |>
  mutate(homeless_flag = case_when(homeless_flag == "total_homeless_enroll" ~ "Homeless",
                                   TRUE ~ "Not Homeless")) |>
  ggplot(aes(reorder(State, percent_homeless, decreasing = TRUE), 
             y = value, 
             fill = fct_rev(homeless_flag))) +
  geom_col() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "State", y = "Total Homeless Enrolled") +
  scale_fill_manual(values = c("Not Homeless" = "grey", "Homeless" = "blue")) +
  geom_text(data = ~ filter(.x, homeless_flag == "Homeless"), 
            aes(label= value), 
            vjust = -0.25)
```
```{r}
percent_homeless_table |>
  mutate(total_not_homeless_enroll = total_child_enroll - total_homeless_enroll) |>
  pivot_longer(cols = c(total_homeless_enroll, total_not_homeless_enroll),
               names_to = "homeless_flag",
               values_to = "value") |>
  mutate(homeless_flag = case_when(homeless_flag == "total_homeless_enroll" ~ "Homeless",
                                   TRUE ~ "Not Homeless")) |>
  ggplot(aes(reorder(State, percent_homeless, decreasing = TRUE), 
             y = value, 
             fill = fct_rev(homeless_flag))) +
  geom_col(position = "fill") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "State", y = "% Homeless Enrolled") +
  scale_fill_manual(values = c("Not Homeless" = "grey", "Homeless" = "blue")) +
  geom_text(data = ~ filter(.x, homeless_flag == "Homeless"), 
            aes(label = round(percent_homeless, 2)),
            position = position_fill(vjust = 0.25))
```

---

## Total Homeless Under Five

```{r}

```

--- 

## Total/% Homeless Under Five Served by Head Start

```{r}

```

---

## Overview of Homeless Counts from Different Sources

```{r}
ed_totals_under_5 |>
  mutate(HOMELESS_3_5_655 = as.numeric(HOMELESS_3_5_655),
         HOMELESS_0_5_818 = as.numeric(HOMELESS_0_5_818)) |>
  select(STATENAME, 
         YEAR_END, 
         HOMELESS_UNDER_5_GRADE_1,
         HOMELESS_3_5_655, 
         HOMELESS_0_5_818, 
         TOTAL_HS_HOMELESS_ENROLL) |>
  pivot_longer(cols = c(HOMELESS_UNDER_5_GRADE_1, 
               HOMELESS_3_5_655, 
               HOMELESS_0_5_818, 
               TOTAL_HS_HOMELESS_ENROLL),
               names_to = "HOMELESS_MEASURE",
               values_to = "HOMELESS_COUNT") |>
  mutate(HOMELESS_MEASURE = case_when(HOMELESS_MEASURE == "HOMELESS_UNDER_5_GRADE_1" ~
                                        "Grade 1 Proxy Method",
                                      HOMELESS_MEASURE == "HOMELESS_3_5_655" ~
                                        "Chidren 3-5 Enrolled in Public Schools (File 655)",
                                      HOMELESS_MEASURE == "HOMELESS_0_5_818" ~
                                        "Children 0-5 Served by McKinney-Vento subgrant funds (File 818)",
                                      HOMELESS_MEASURE == "TOTAL_HS_HOMELESS_ENROLL" ~
                                        "Homeless Children Enrolled in Head Start")) |>
  filter(STATENAME == "MASSACHUSETTS") |>
  ggplot(aes(x = YEAR_END, y = HOMELESS_COUNT, col = HOMELESS_MEASURE, group = HOMELESS_MEASURE)) +
  geom_line() +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 5))  +
  geom_text_repel(aes(label = HOMELESS_COUNT), 
            vjust = -0.5,
            size = 3) +
  labs(x = "Year", y = "Homeless Count")
  
```


---

> Shelter type is not reported disaggregated by grade. We took the percentage of K-12 students in each shelter type, each year for each state and multiplied that percentage by the number of homeless children under 5 identified via the Grade 1 proxy method

```{r}
ed_totals_under_5_char |>
  filter(STATENAME == "MASSACHUSETTS") |>
  ggplot(aes(x = YEAR_END, y = TOTAL_CHAR_UNDER_5, fill = Characteristics)) +
  geom_col(position = "dodge", width = 0.9) +
  theme_minimal() +
  geom_text(aes(label = round(TOTAL_CHAR_UNDER_5, 0)),
            position = position_dodge(width = 0.9),
            size = 3,
            vjust = -0.25  )
```
```{r}
ed_totals_under_5_char |>
  filter(STATENAME == "MASSACHUSETTS") |>
  ggplot(aes(x = YEAR_END, y = TOTAL_CHAR_UNDER_5, fill = Characteristics)) +
  geom_col(position = "fill") +
  theme_minimal() +
  geom_text(aes(label = paste(as.character(round(PERCENT_CHAR_ST_YR, 4) * 100) , "%")),
            position = position_fill(),
            size = 3,
            vjust = -0.25  )
```



